---
title: "summary of aquaria reads"
author: "Kimberly Ledger"
date: "2023-04-27"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

this code summarizes the aquaria results after doing some initial test for filter in the code "gadid_ASV_analysis_20230421_aquaria.Rmd" and testing for differences across treatments in the code "gadid_aquaria_20230421_stats.Rmd".

load libraries
```{r}
library(tidyverse)
library(ggplot2)
library(forcats)
library(cowplot)
```


load data 
```{r}
aquaria_dat <- read.csv("/genetics/edna/workdir/gadids/20230421_aquaria/20230421_filtered_readsummary.csv") %>%
  dplyr::select(!X) %>%
  dplyr::select(!extraction_plate) %>%
  dplyr::select(!extraction_well) %>%
  rename("prep" = "replicate")  ##rename replicate to not confuse as true replicates, these are the same sample but used different library prep methods
```


since we've learned that the prep method and the extraction replicate do not significantly affect the composition of samples, i will treat the different Sample_IDs within an alternative_ID as replicates. 

```{r echo=FALSE}
aquaria_dat %>% 
  filter(sample_type == "sample") %>%
  group_by(Sample_ID) %>%
  mutate(sum=sum(reads)) %>%
  mutate(prop = reads/sum) %>%
  ggplot(aes(x=Sample_ID, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  facet_wrap(~alternative_ID, scales = 'free_x') + 
  theme_bw() +
  labs(
    y = "proportion of sequencing reads",
    x = "sample",
    title = "aquaria - proportion of reads") + 
  theme(
    axis.text.x = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )
```


my guess is that any alt_IDs with 4 or more 0's out of the six replicates was probably a "field blank". let's make a list of those. 

```{r}
possible_field_blanks <- aquaria_dat %>% 
  filter(sample_type == "sample") %>%
  group_by(alternative_ID, Sample_ID) %>%
  mutate(sum=sum(reads)) %>%
  mutate(prop = reads/sum) %>%
  summarise(count = sum(sum == 0)/4) %>%
  group_by(alternative_ID) %>%
  summarize(num_of_zeros = sum(count)) %>%
  filter(num_of_zeros >= 4)
possible_field_blanks
```

next i will remove the samples i think might be field blanks, and then i will average the compostion in the replicates and plot. just averaging for now to make a visual. likely want to retain replicates in actual analyses (?)


```{r}
blanks <- possible_field_blanks$alternative_ID

sample_avg <- aquaria_dat %>% 
  filter(sample_type == "sample") %>%
  filter(alternative_ID != 107) %>%  #for some reason filtering by "blanks" didn't work so i'm going the long way... 
  filter(alternative_ID != 118) %>%
  filter(alternative_ID != 178) %>%
  filter(alternative_ID != 194) %>%
  filter(alternative_ID != 229) %>%
  filter(alternative_ID != 241) %>%
  filter(alternative_ID != 385) %>%
  filter(alternative_ID != 395) %>%
  filter(alternative_ID != 430) %>%
  filter(alternative_ID != 52) %>%
  group_by(Sample_ID) %>%
  mutate(sum=sum(reads)) %>%
  mutate(prop = reads/sum) %>%  
  group_by(alternative_ID, taxon) %>%
  summarize(avg_prop = mean(prop, na.rm = TRUE),
            sd_prop = sd(prop))
```

to change the order of the stacked barplot, i need to redo the format
```{r}
sample_avg_wide <- sample_avg %>%
  dplyr::select(!sd_prop) %>%
  pivot_wider(names_from = "taxon", values_from = "avg_prop") %>%
  dplyr::arrange(desc(`Gadus macrocephalus`))

my_order <- as.character(sample_avg_wide$alternative_ID)

library(forcats)

sample_avg_wide_test <- sample_avg_wide %>%
  mutate(alternative_ID = factor(alternative_ID)) %>%
  mutate(alternative_ID = fct_relevel(alternative_ID, my_order))

sample_avg_long <- sample_avg_wide_test %>%
  pivot_longer(cols = c(2:5), names_to = "taxon", values_to = "avg_prop")

sample_avg_long %>%
  ggplot(aes(x=alternative_ID, y=avg_prop, fill=taxon)) +
  geom_bar(stat = "identity", position = "stack") + 
  theme_bw() +
  labs(
    y = "proportion of sequencing reads",
    x = "sample",
    title = "aquaria - proportion of reads - pcod descending") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )
```

```{r}
sample_avg_wide <- sample_avg %>%
  dplyr::select(!sd_prop) %>%
  pivot_wider(names_from = "taxon", values_from = "avg_prop") %>%
  dplyr::arrange(`Gadus chalcogrammus`)

my_order <- as.character(sample_avg_wide$alternative_ID)

library(forcats)

sample_avg_wide_test <- sample_avg_wide %>%
  mutate(alternative_ID = factor(alternative_ID)) %>%
  mutate(alternative_ID = fct_relevel(alternative_ID, my_order))

sample_avg_long <- sample_avg_wide_test %>%
  pivot_longer(cols = c(2:5), names_to = "taxon", values_to = "avg_prop")

sample_avg_long %>%
  ggplot(aes(x=alternative_ID, y=avg_prop, fill=taxon)) +
  geom_bar(stat = "identity", position = "stack") + 
  theme_bw() +
  labs(
    y = "proportion of sequencing reads",
    x = "sample",
    title = "aquaria - proportion of reads - pollock ascending") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )
```

```{r}
sample_avg_wide <- sample_avg %>%
  dplyr::select(!sd_prop) %>%
  pivot_wider(names_from = "taxon", values_from = "avg_prop") %>%
  dplyr::arrange(`Boreogadus saida`)

my_order <- as.character(sample_avg_wide$alternative_ID)

library(forcats)

sample_avg_wide_test <- sample_avg_wide %>%
  mutate(alternative_ID = factor(alternative_ID)) %>%
  mutate(alternative_ID = fct_relevel(alternative_ID, my_order))

sample_avg_long <- sample_avg_wide_test %>%
  pivot_longer(cols = c(2:5), names_to = "taxon", values_to = "avg_prop")

sample_avg_long %>%
  ggplot(aes(x=alternative_ID, y=avg_prop, fill=taxon)) +
  geom_bar(stat = "identity", position = "stack") + 
  theme_bw() +
  labs(
    y = "proportion of sequencing reads",
    x = "sample",
    title = "aquaria - proportion of reads - arctic cod ascending") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )
```

modify plot to group/facet(?) samples by the dominate species 
```{r}
sample_avg_wide <- sample_avg %>%
  dplyr::select(!sd_prop) %>%
  pivot_wider(names_from = "taxon", values_from = "avg_prop") 

samp_id <- sample_avg_wide$alternative_ID

sample_avg_test <- sample_avg_wide[,-1] %>%
  mutate(group_max = names(.)[max.col(.)]) %>%
  mutate(alternative_ID = samp_id)

sample_avg_group <- sample_avg_test %>% 
    pivot_longer(cols = c(1:4), names_to = "taxon", values_to = "avg_prop")
```


```{r}
sample_avg_group %>%
  ggplot(aes(x=alternative_ID, y=avg_prop, fill=taxon)) +
  geom_bar(stat = "identity", position = "stack") + 
  theme_bw() +
  facet_grid(~group_max, scales = 'free_x') +
  labs(
    y = "proportion of sequencing reads",
    x = "sample",
    title = "aquaria - proportion of reads") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )
```

okay, i don't know how i would arrange the contents of each facet by different species, so i'll try dividing into three different plots and then combining... 

```{r}
bs1 <- sample_avg_group %>%
  filter(group_max == "Boreogadus saida") %>%
  pivot_wider(names_from = "taxon", values_from = "avg_prop") %>%
  dplyr::arrange(desc(`Boreogadus saida`))

my_order <- as.character(bs1$alternative_ID)

bs2 <- bs1 %>%
  mutate(alternative_ID = factor(alternative_ID)) %>%
  mutate(alternative_ID = fct_relevel(alternative_ID, my_order))

bs3 <- bs2 %>%
  pivot_longer(cols = c(3:6), names_to = "taxon", values_to = "avg_prop") 
  
bs_plot <- bs3 %>%  
  ggplot(aes(x=alternative_ID, y=avg_prop, fill=taxon)) +
  geom_bar(stat = "identity", position = "stack") + 
  theme_bw() +
  labs(
    y = "proportion of sequencing reads",
    x = "sample")+ #,
    #title = "arctic cod dominate") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "none",
    legend.title = element_blank()
  )
bs_plot
```

```{r}
gc1 <- sample_avg_group %>%
  filter(group_max == "Gadus chalcogrammus") %>%
  pivot_wider(names_from = "taxon", values_from = "avg_prop") %>%
  dplyr::arrange(desc(`Gadus chalcogrammus`))

my_order <- as.character(gc1$alternative_ID)

gc2 <- gc1 %>%
  mutate(alternative_ID = factor(alternative_ID)) %>%
  mutate(alternative_ID = fct_relevel(alternative_ID, my_order))

gc3 <- gc2 %>%
  pivot_longer(cols = c(3:6), names_to = "taxon", values_to = "avg_prop") 
  
gc_plot <- gc3 %>%  
  ggplot(aes(x=alternative_ID, y=avg_prop, fill=taxon)) +
  geom_bar(stat = "identity", position = "stack") + 
  theme_bw() +
  labs(
    y = "proportion of sequencing reads",
    x = "sample") + #,
    #title = "pollock dominate") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "none",
    legend.title = element_blank()
  )
gc_plot
```

```{r}
gm1 <- sample_avg_group %>%
  filter(group_max == "Gadus macrocephalus") %>%
  pivot_wider(names_from = "taxon", values_from = "avg_prop") %>%
  dplyr::arrange(desc(`Gadus macrocephalus`))

my_order <- as.character(gm1$alternative_ID)

gm2 <- gm1 %>%
  mutate(alternative_ID = factor(alternative_ID)) %>%
  mutate(alternative_ID = fct_relevel(alternative_ID, my_order))

gm3 <- gm2 %>%
  pivot_longer(cols = c(3:6), names_to = "taxon", values_to = "avg_prop") 
  
gm_plot <- gm3 %>%  
  ggplot(aes(x=alternative_ID, y=avg_prop, fill=taxon)) +
  geom_bar(stat = "identity", position = "stack") + 
  theme_bw() +
  labs(
    y = "proportion of sequencing reads",
    x = "sample") + # ,
    #title = "pacific cod dominate") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "none",
    legend.title = element_blank()
  )
gm_plot
```
save legend
```{r}
legend <- cowplot::get_legend(bs_plot)
```


```{r proportion_fig, fig.height=5, fig.width=6}
plot_grid(gm_plot, bs_plot, gc_plot, legend, labels = c("A", "B", "C", ""), ncol = 2, rel_widths = c(1.5,1))
```





think of a way to include errors in proportions into a visual...? 


next it would be great to have some metadata for these samples to know replicates and species proportions 

