---
title: "gadid field application"
author: "Kimberly Ledger"
date: "2023-09-27"
output: html_document
---

this script uses gadid metabarcoding reads from the DBO and GOApcod samples

load libraries
```{r, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(ggOceanMaps)
library(ggspatial)
```

load data
```{r}
## DBO metadata - includes ALL samples
DBO_metadata <- read.csv("/home/kimberly.ledger/gadid_metabarcoding/gadid_aquariaDBO_metadata.csv") %>%
  filter(project == "DBO")
#illumina output changed "_" to "-"
DBO_metadata$Sample_ID <- gsub("_", "-", DBO_metadata$Sample_ID)

DBO_metadata$station_ID <- as.factor(DBO_metadata$station_ID)
DBO_metadata$station <- as.factor(DBO_metadata$station)
DBO_metadata$cast <- as.factor(DBO_metadata$cast)

## DBO gadid reads - these have been decontaminated and zero read samples removed
DBO_data <- read.csv("~/gadid_metabarcoding/gadid_fieldapplication/DBO_taxonreads_20230927.csv") %>%
  select(!X)

DBO_data$taxon <- as.factor(DBO_data$taxon)
```

start by plotting all of the DBO sampling locations on a map 
```{r}
DBO_sites <- DBO_metadata %>%
  filter(sample_type == "sample") %>%
  select(location1, station, longitude, latitude) %>%
  unique() %>%
  filter(!is.na(longitude))
```

add sites to map 
```{r}
map_with_sites <- basemap(c(-177, -159, 61, 73), bathymetry = FALSE, rotate = TRUE) + 
  ggspatial::geom_spatial_point(data = DBO_sites, aes(x = longitude, y = latitude, color = station), size = 1) + 
  labs(color = "Station")
  

#  theme(legend.key = element_rect(fill = "transparent"))

map_with_sites
```

okay, where 1 and 2 are Bering Sea and 3 and 4 are Chukchi Sea. maybe i want to adjust colors, later. 


now work with the read data  ---- the metadata associated with these reads has NOT been updated (as of 20230927) so i will just extract read info and join to the metadata loaded here. 
also, i want to keep ALL the samples, even the ones with zero reads 
```{r}
DBO_metadata_mini <- DBO_metadata %>%
  filter(sample_type == "sample") %>%
  select(Sample_ID, station_ID, location1, station, cast, longitude, latitude)

DBO_data_mini <- DBO_data %>%
  select(Sample_ID, taxon, total_reads)

DBO_all <- DBO_metadata_mini %>%
  left_join(DBO_data_mini)
```
```{r}
my_palette <- brewer.pal(3, "BuPu")
```

plot transect 1 Bering Sea samples   
```{r}
DBO_all %>%
  #filter(location1 == "Bering Sea) %>%
  filter(station == "1") %>%
  group_by(Sample_ID) %>%
  mutate(sum=sum(total_reads)) %>%
  mutate(prop = total_reads/sum) %>%
  ggplot(aes(x=Sample_ID, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = my_palette[2:3]) +
  facet_wrap(~cast, scales = 'free', ncol = 3) +
  theme_bw() + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    axis.text.x = element_blank(),
    legend.position = "right",
    legend.title = element_blank()) + 
  labs(title = "Bering Sea - Station 1")
```

plot transect 2 Bering Sea samples   
```{r}
DBO_all %>%
  #filter(location1 == "Bering Sea) %>%
  filter(station == "2") %>%
  group_by(Sample_ID) %>%
  mutate(sum=sum(total_reads)) %>%
  mutate(prop = total_reads/sum) %>%
  ggplot(aes(x=Sample_ID, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = my_palette[2:3]) +
  facet_wrap(~cast, scales = 'free', ncol = 3) +
  theme_bw() + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    axis.text.x = element_blank(),
    legend.position = "right",
    legend.title = element_blank()) + 
  labs(title = "Bering Sea - Station 2")
```

plot transect 3 Chukchi Sea samples  
```{r}
DBO_all %>%
  #filter(location1 == "Chukchi Sea") %>%
  filter(station == 3) %>%
  group_by(Sample_ID) %>%
  mutate(sum=sum(total_reads)) %>%
  mutate(prop = total_reads/sum) %>%
  ggplot(aes(x=Sample_ID, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = my_palette[2]) +
  facet_wrap(~cast, scales = 'free', ncol = 3) +
  theme_bw() + 
  theme(
    axis.text.x = element_blank(),
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "right",
    legend.title = element_blank()) + 
  labs(title = "Chukchi Sea - Station 3")
```


plot transect 4 Chukchi Sea samples  
```{r}
DBO_all %>%
  #filter(location1 == "Chukchi Sea") %>%
  filter(station == 4) %>%
  group_by(Sample_ID) %>%
  mutate(sum=sum(total_reads)) %>%
  mutate(prop = total_reads/sum) %>%
  ggplot(aes(x=Sample_ID, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = my_palette) +
  facet_wrap(~cast, scales = 'free', ncol = 3) +
  theme_bw() + 
  theme(
    axis.text.x = element_blank(),
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "right",
    legend.title = element_blank()) + 
  labs(title = "Chukchi Sea - Station 4")
```

plot M8
```{r}
DBO_all %>%
  filter(station_ID == "M8")%>%
  group_by(Sample_ID) %>%
  mutate(sum=sum(total_reads)) %>%
  mutate(prop = total_reads/sum) %>%
  ggplot(aes(x=Sample_ID, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = my_palette[2]) +
  #facet_wrap(~cast, scales = 'free', ncol = 3) +
  theme_bw() + 
  theme(
    axis.text.x = element_blank(),
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "right",
    legend.title = element_blank()) + 
  labs(title = "M8")
```
plot C2
```{r}
DBO_all %>%
  filter(station_ID == "C2")%>%
  group_by(Sample_ID) %>%
  mutate(sum=sum(total_reads)) %>%
  mutate(prop = total_reads/sum) %>%
  ggplot(aes(x=Sample_ID, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = my_palette[2]) +
  #facet_wrap(~cast, scales = 'free', ncol = 3) +
  theme_bw() + 
  theme(
    axis.text.x = element_blank(),
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "right",
    legend.title = element_blank()) + 
  labs(title = "C2")
```


####### Now i want to dive into the GOApcod samples i did gadid metabarcoding with way back when... 
```{r}

```



